Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 method.tex
pass after rearrangement
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@952 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the \CBC{} waveform
space that reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs.  We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our novel detection statistic using the time-slice
decomposition and the \SVD.

@@

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
\FIXME{This quantity is related to the early warning \SNR{} by the normalization
of the partial filter defined by the present time slice.}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multirate filter network with outputs for
each of the early warning \SNR{}s.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional \TD{} or \FD{} matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in decimated time slice $s$\\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

>>>>

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method described in~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early warning detection statistic using the time-slice
decomposition and the \SVD.

@@

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multirate filter network with outputs for
each of the early warning \SNR{}s.

@@

We now examine the computational cost scaling of the conventional \TD{} or
\FD{} matched filter procedure as compared with \lloid{}.  For convenience,
table~\ref{tab:recap} provides a review of the notation that we will need in
this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in decimated time slice $s$\\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}
